Inference using context-dependent models of mutation via bounds on information propagation

### 2021-12-11 writeups/context-dependence.tex
edits for EMs

%%%%%%%%
\subsection{Model fit and model selection}
%EM I promoted this section here because we use residuals a lot in the results.
Fitting a model to nucleotide data will likely require iterative modification of the model.
The difference between observed and expected T-mer counts provide a good assessment of how and whether a model can be improved.
%EM Why shorter? Because there are fewer of them?
These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model.
%EM Should we replace a and b with x and y here to agree with the rest of the doc?
For each T-mer $(a,b)$, write $O_{a,b}$ for the number of this T-mer observed in the data,
and $E_{a,b}$ for the number expected under the model.
(The expected number is the number of occurrences of $a$ in the initial sequence
multiplied by the probability $p_{n,\ell,r}(t;a,b)$, where $n$ is the length of $a$,
and $\ell$, $r$ are the appropriate overhangs.)
If the T-mers were nonoverlapping, this would form a standard contingency table,
and standard practice would be to divide the residuals by the square root of the expected counts
and compare these to a standard Normal distribution.
However, there are are $n$ possible such tables --
so, we compute normalized residuals as
\begin{align}
    Z_{a,b} = \frac{ O_{a,b} - E_{a,b} }{ \sqrt{n E_{a,b}} },
\end{align}
which is equivalent to averaging the $z$-scores of the $n$ possible tables of nonoverlapping counts.
These can then be compared to a standard Normal distribution.

@@

To begin, we fit an unconstrained model of single-nucleotide substitution using (5,1,1) T-mers
(three sites at the base with an overhang of one on each side)
using a constrained quasi-Newton method (\verb|optim(.., method='L-BFGS-B')| in R).

\begin{table}[ht]
  \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }ll}
          \hline
            x & y & truth & MLE \\
          \hline
          %EM I'm confused why we have MLE parameters for these single NT terms because I don't see those being used in the iterative model fitting. I thought we started with a (5,1,1) model.
          \nA  &   \nT           & 0.10 & 0.0996 \\
          \nT  &   \nA           & 0.10 & 0.0982 \\
          \nC  &   \nG           & 0.15 & 0.1505 \\
          \nG  &   \nC           & 0.15 & 0.1497 \\
          \nA  &   \nC           & 0.08 & 0.0784 \\
          \nT  &   \nG           & 0.08 & 0.0792 \\
          \nA  &   \nG           & 0.08 & 0.0798 \\
          \nT  &   \nC           & 0.08 & 0.0795 \\
          \nC  &   \nA           & 0.12 & 0.1202 \\
          \nG  &   \nT           & 0.12 & 0.1196 \\
          \nC  &   \nT           & 0.12 & 0.1198 \\
          \nG  &   \nA           & 0.12 & 0.1193 \\
       \nC\nG  &  \nT\nG         & 0.40 & 0.3978 \\
       \nC\nG  &  \nC\nA         & 0.40 & 0.3978 \\
    \nT\nC\nC  &  \nT\nT\nC      & 0.10 & 0.1033 \\
    \nA\nG\nG  &  \nA\nA\nG      & 0.10 & 0.1033 \\
           \hline
        \end{tabular}
  \end{center}
  \caption{
      True parameter values, and MLE estimates obtained
      from iterative fitting as described in the text to a $10^6$ bp sequence
    evolved to have 28.5\% sequence divergence
    with the above mutational motifs. The transitions $\nC\nG \to \nT\nG$ and $\nC\nG \to \nC\nA$
    were constrained during fitting to be equal (DNA strand symmetry),
    as were the transitions $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$.
    \label{tab:cpg_results} }
\end{table}

We find that the (2,1,0) and (2,0,1) T-mer residuals of this model are dominated by
terms involving CG (Table~\ref{tab:resid1}).
%EM Stupid question: why weren't these picked up by the 5,1,1 model?
%EM Why isn't the starting point a (1,0,0) model?
These results clearly call for addition of the CpG mutation motif.

\begin{table}
%EM This table is BIG. And it seems like our conclusions only stem from the very biggest residuals.
%EM What if we cut of anything that doesn't have an absolute Z-score bigger than 10?
% cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l0.tsv
% cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l1.tsv
    \begin{center}
        \begin{tabular}{ccrrrr}
                $x$ & $y$ & observed &   expected &    residual &  $Z$ \\
                \hline
                \nC\nG  & \nC\_   & 100410   &  120303    & -19892.50   &	-40.55  \\
                \nC\nA  & \nT\_   &  19884   &   28350    &  -8466.43   &	-35.55  \\
                \nC\nT  & \nT\_   &  19998   &   28061    &  -8063.46   &	-34.03  \\
                \nC\nC  & \nT\_   &  23673   &   28177    &  -4503.96   &	-18.97  \\
                \nG\nG  & \nC\_   &  17742   &   19418    &  -1675.92   &	 -8.50  \\
                \nC\nG  & \nG\_   &  18450   &   19478    &  -1028.34   &	 -5.21  \\
                \nA\nG  & \nC\_   &  10968   &   11760    &   -791.73   &	 -5.16  \\
                %EM It looks like this is missing an entry?
                \nG\nG  & \nG\_   & 118805   &  120975    &  -2170.25   &    \\
                \hline
                \nA\nG  & \nT\_   &  17280   &   16515    &    765.20   &  4.21   \\
                \nG\nT  & \nC\_   &  20490   &   19507    &    983.44   &  4.97   \\
                \nG\nG  & \nT\_   &  20929   &   19278    &   1650.83   &  8.40   \\
                \nC\nC  & \nC\_   & 124149   &  119880    &   4268.52   &  8.71   \\
                \nG\nG  & \nA\_   &  30483   &   28288    &   2195.34   &  9.22   \\
                \nC\nT  & \nC\_   & 126957   &  119389    &   7567.92   & 15.48   \\
                \nC\nA  & \nC\_   & 128673   &  120619    &   8054.46   & 16.39   \\
                \nC\nG  & \nT\_   &  49311   &   28276    &  21034.85   & 88.45   \\
                \hline
                \hline
                \nC\nG  &  \_\nG  &  100803  &  120541  &  -19737.80  &  -40.19 \\
                \nT\nG  &  \_\nA  &   20040  &   28179  &   -8139.45  &  -34.28 \\
                \nG\nG  &  \_\nA  &   20985  &   28288  &   -7302.96  &  -30.70 \\
                \nA\nG  &  \_\nA  &   22929  &   28399  &   -5469.73  &  -22.95 \\
                \nC\nC  &  \_\nG  &   17805  &   19410  &   -1605.01  &   -8.14 \\
                \nC\nT  &  \_\nG  &   10470  &   11684  &   -1214.31  &   -7.94 \\
                \nC\nA  &  \_\nG  &   10839  &   11882  &   -1043.39  &   -6.76 \\
                \nC\nG  &  \_\nC  &   18354  &   19348  &    -994.19  &   -5.05 \\
                \hline
                \nC\nT  &  \_\nA  &   17007  &   16078  &     929.33  &    5.18 \\
                \nT\nC  &  \_\nG  &   20691  &   19509  &    1182.40  &    5.98 \\
                \nC\nC  &  \_\nA  &   20640  &   19160  &    1480.44  &    7.56 \\
                \nT\nC  &  \_\nT  &   30309  &   28320  &    1988.91  &    8.35 \\
                \nA\nG  &  \_\nG  &  127191  &  121450  &    5740.76  &   11.64 \\
                \nG\nG  &  \_\nG  &  127946  &  120977  &    6969.46  &   14.16 \\
                \nT\nG  &  \_\nG  &  127539  &  120512  &    7026.52  &   14.31 \\
                \nC\nG  &  \_\nA  &   49098  &   28186  &   20911.93  &   88.07 \\
                \hline
        \end{tabular}
    \end{center}
    \caption{
        Largest residuals of shape (2,1,0) and (2,0,1) after fitting a model with only ``basic''
        (single-nucleotide) mutations to sequence simulated under the model
        of Table~\ref{tab:cpg_results}, that also includes CpG hypermutability
        as well as $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$ mutations.
        Note that in the table, the underscore means ``any nucleotide''.
        \label{tab:resid1}
    }
\end{table}

%EM "These patterns": Is it exactly the patterns in the true model?
%EM Would we need to have prior information about CpGs to pick those additional terms?
After adding these patterns into the model and re-fitting, we searched for evidence of additional processes happening in the data.
% starting at the parameters inferred from the first run and a CpG rate of 0.0,
%EM what is "statistically significant" here? I searched around and found 1.96 corresponding to 0.05.
To do so, we calculated length 2 residuals and (3,1,1) T-mer residuals.
Although some of the length 2 residuals were statistically significant,
none were as large as those seen in (3,1,1) T-mer residuals, which are shown in Table~\ref{tab:resid2}.
These residuals clearly call for adding the mutation tuples $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$.
These were indeed the motifs used when simulating the data.
After adding these in,
% and initializing all parameters to 0.1
no further (3,1,1) T-mer residuals achieved
statistical significance after a Bonferroni correction;
the final parameter values are shown in Table \ref{tab:cpg_results};
parameter estimates differ from the truth by less than 3\% (mostly within 1\%).

@@

\begin{figure}
    \begin{center}
        %EM Should I be looking at the gray bars or black bars? What are they? I'm guessing 95% is gray, because that corresponds to good coverage.
        \includegraphics{writeup-plots/coverage_results}
    \end{center}
    \caption{
        Credible intervals for the four parameters of the dynamic Ising model,
        calculated using $(6,2,2)$ T-mers,
        from 61 MCMC runs on independently simulated sequences of $10^6$ bases
        differing at around 18.6\% of the sites.
        Shown are relative intervals, obtained by dividing by the true value.
        \label{fig:ising_coverage}}
\end{figure}

@@

Nevertheless, we find significant effects for a number of biochemically-informed mutation features (Figure~\ref{fig:biochem_results}).
Specifically, we find strong CpG and UV damage effects.
%EM Kelley saw the same thing https://paperpile.com/shared/aAYSgW ...
It may seem tenuous to look for mutational signatures associated with cancer in the germline,
or absurd that UV damage could play a significant role.
This is in part due to the ``streetlight effect'' --
we can't look for signatures that are not characterized --
but it is also reasonable that shared repair machinery might induce similar context-dependent errors.
More likely sources of error in the germline include the effects of recombination
\citep{arbeithuber2015crossovers,myers2010drive}.

@@

%EM given the full likelihood arguments, can we really say provably correct?
This paper presents a tractable, provably correct method
for doing statistical inference using observations of a particle model
in which probabilities of change are affected by nearby sites.
Such models appear in many fields, for instance,
%EM are semicolons really legal here?
as the dynamic counterparts to Markov random fields;
many spin glass models;
and models of DNA sequence evolution.

%EM approximate full likelihoods?
The methods described here provide a way to efficiently calculate full likelihoods,
which in many circumstances provide the optimal basis for inference \citep{neyman1933problem},
fast enough to allow Bayesian inference from Monte Carlo sampling.
The method is fast thanks to its dependence on only local neighborhoods
(so the time does not scale with sequence length once T-mer abundances are counted)
and its reliance on sparse matrices whose structure can be precomputed and cached.
It avoids approximations made by previous methods thanks to the use of T-mers
and explicit quantification of the necessary length of the ``arms'' of the T-mers.
We first discuss possible uses of the method,
followed by challenges (and possible solutions).

@@

\paragraph{Limitations}
The model presented assumes that the same process occurs at every location along the sequence,
and that the length of the sequence does not change (no insertions or deletions).
Neither of these assumptions are likely to be true for any data set of DNA sequence;
however, both assumptions are also commonly made in phylogenetics.
Some relaxation of the assumptions is possible -- for instance,
we can easily allow heterogeneity between sites by scaling rates by an overall gamma-distributed random factor \citep{yang1994maximum}.
%EM Does a gap character really make sense? It seems like this changes the interpretation of the tmer length... I'd skip.
Short deletions might be accommodated by introducing a ``gap'' letter to the alphabet,
but insertions are problematic.

%EM Full likelihood conjecture?

>>>>

%%%%%%%%
\subsection{Model fit and model selection}
Fitting a model to nucleotide data will likely require iterative modification of the model.
The difference between observed and expected T-mer counts provide a good assessment of how and whether a model can be improved.
These \emph{residual} counts are usually most usefully computed for T-mers shorter than those used to fit the model,
because the smaller total number of possible T-mers makes it easier to identify the shared patterns.
For each T-mer $(x,y)$, write $O_{x,y}$ for the number of this T-mer observed in the data,
and $E_{x,y}$ for the number expected under the model.
(The expected number is the number of occurrences of $x$ in the initial sequence
multiplied by the probability $p_{n,\ell,r}(t;x,y)$, where $n$ is the length of $x$,
and $\ell$, $r$ are the appropriate overhangs.)
If the T-mers were nonoverlapping, this would form a standard contingency table,
and standard practice would be to divide the residuals by the square root of the expected counts
and compare these to a standard Normal distribution.
However, there are are $n$ possible such tables --
so, we compute normalized residuals as
\begin{align}
    Z_{x,y} = \frac{ O_{x,y} - E_{x,y} }{ \sqrt{n E_{x,y}} },
\end{align}
which is equivalent to averaging the $z$-scores of the $n$ possible tables of nonoverlapping counts.
These can then be compared to a standard Normal distribution.

@@

To begin, we fit an unconstrained model of single-nucleotide substitution
using a constrained quasi-Newton method (\verb|optim(.., method='L-BFGS-B')| in R).
Recall that for inference, we next need to pick a T-mer shape
with which to compute likelihoods,
and the shape of this T-mer is only determined by the shape of dependencies in the generative model
in that it determines the minimum base and overhang lengths.
We use (5,1,1) T-mers: three sites at the base with an overhang of one on each side.
Since this initial model has no context-dependence,
we could do the first step of model fitting using (1,0,0) T-mers (i.e., single positions),
but for simplicity we use the same T-mer shape for all steps.

After fitting the model, we can look at residuals using any T-mer shape we like.
At this first step, we computed the (2,1,0) and (2,0,1) T-mer residuals,
shown in Table~\ref{tab:resid1}.
The largest residuals are an excess of $\nC \to \nT$ mutations in $\nC\nG$ dinucleotides,
and the reverse-complement pattern ($z$ scores of 88).
Complementing this are deficits of $\nC \to \nT$ mutations in other contexts,
since the $\nC \to \nT$ mutation rate has been fit to be an average across contexts.
This clearly calls for addition of the $\nC\nG \to \nT\nG$ mutation motif (and reverse complement),
which was used in simulating the data.

\begin{table}[ht]
  \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }ll}
          \hline
            x & y & truth & MLE \\
          \hline
          \nA  &   \nT           & 0.10 & 0.0996 \\
          \nT  &   \nA           & 0.10 & 0.0982 \\
          \nC  &   \nG           & 0.15 & 0.1505 \\
          \nG  &   \nC           & 0.15 & 0.1497 \\
          \nA  &   \nC           & 0.08 & 0.0784 \\
          \nT  &   \nG           & 0.08 & 0.0792 \\
          \nA  &   \nG           & 0.08 & 0.0798 \\
          \nT  &   \nC           & 0.08 & 0.0795 \\
          \nC  &   \nA           & 0.12 & 0.1202 \\
          \nG  &   \nT           & 0.12 & 0.1196 \\
          \nC  &   \nT           & 0.12 & 0.1198 \\
          \nG  &   \nA           & 0.12 & 0.1193 \\
       \nC\nG  &  \nT\nG         & 0.40 & 0.3978 \\
       \nC\nG  &  \nC\nA         & 0.40 & 0.3978 \\
    \nT\nC\nC  &  \nT\nT\nC      & 0.10 & 0.1033 \\
    \nA\nG\nG  &  \nA\nA\nG      & 0.10 & 0.1033 \\
           \hline
        \end{tabular}
  \end{center}
  \caption{
      True parameter values, and MLE estimates obtained
      from iterative fitting as described in the text to a $10^6$ bp sequence
    evolved to have 28.5\% sequence divergence
    with the above mutational motifs. The transitions $\nC\nG \to \nT\nG$ and $\nC\nG \to \nC\nA$
    were constrained during fitting to be equal (DNA strand symmetry),
    as were the transitions $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$.
    \label{tab:cpg_results} }
\end{table}

\begin{table}
%EM This table is BIG. And it seems like our conclusions only stem from the very biggest residuals.
%EM What if we cut of anything that doesn't have an absolute Z-score bigger than 10?
%PR True, but I'd like to show that the $z$-scores do down to something smaller?
%PR I've cut out four lines...
% cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l0.tsv
% cpg-plus-epsilon/32140/base-model-fit-5-3-l1-resids.2.1.l1.tsv
    \begin{center}
        \begin{tabular}{ccrrrr}
                $x$ & $y$ & observed &   expected &    residual &  $Z$ \\
                \hline
                \nC\nG  & \nC\_   & 100410   &  120303    & -19892.50   &	-40.55  \\
                \nC\nA  & \nT\_   &  19884   &   28350    &  -8466.43   &	-35.55  \\
                \nC\nT  & \nT\_   &  19998   &   28061    &  -8063.46   &	-34.03  \\
                \nC\nC  & \nT\_   &  23673   &   28177    &  -4503.96   &	-18.97  \\
                \nG\nG  & \nC\_   &  17742   &   19418    &  -1675.92   &	 -8.50  \\
                \nC\nG  & \nG\_   &  18450   &   19478    &  -1028.34   &	 -5.21  \\
                \nA\nG  & \nC\_   &  10968   &   11760    &   -791.73   &	 -5.16  \\
                \hline
                \nG\nT  & \nC\_   &  20490   &   19507    &    983.44   &  4.97   \\
                \nG\nG  & \nT\_   &  20929   &   19278    &   1650.83   &  8.40   \\
                \nC\nC  & \nC\_   & 124149   &  119880    &   4268.52   &  8.71   \\
                \nG\nG  & \nA\_   &  30483   &   28288    &   2195.34   &  9.22   \\
                \nC\nT  & \nC\_   & 126957   &  119389    &   7567.92   & 15.48   \\
                \nC\nA  & \nC\_   & 128673   &  120619    &   8054.46   & 16.39   \\
                \nC\nG  & \nT\_   &  49311   &   28276    &  21034.85   & 88.45   \\
                \hline
                \hline
                \nC\nG  &  \_\nG  &  100803  &  120541  &  -19737.80  &  -40.19 \\
                \nT\nG  &  \_\nA  &   20040  &   28179  &   -8139.45  &  -34.28 \\
                \nG\nG  &  \_\nA  &   20985  &   28288  &   -7302.96  &  -30.70 \\
                \nA\nG  &  \_\nA  &   22929  &   28399  &   -5469.73  &  -22.95 \\
                \nC\nC  &  \_\nG  &   17805  &   19410  &   -1605.01  &   -8.14 \\
                \nC\nT  &  \_\nG  &   10470  &   11684  &   -1214.31  &   -7.94 \\
                \nC\nA  &  \_\nG  &   10839  &   11882  &   -1043.39  &   -6.76 \\
                \hline
                \nT\nC  &  \_\nG  &   20691  &   19509  &    1182.40  &    5.98 \\
                \nC\nC  &  \_\nA  &   20640  &   19160  &    1480.44  &    7.56 \\
                \nT\nC  &  \_\nT  &   30309  &   28320  &    1988.91  &    8.35 \\
                \nA\nG  &  \_\nG  &  127191  &  121450  &    5740.76  &   11.64 \\
                \nG\nG  &  \_\nG  &  127946  &  120977  &    6969.46  &   14.16 \\
                \nT\nG  &  \_\nG  &  127539  &  120512  &    7026.52  &   14.31 \\
                \nC\nG  &  \_\nA  &   49098  &   28186  &   20911.93  &   88.07 \\
                \hline
        \end{tabular}
    \end{center}
    \caption{
        Largest residuals of shape (2,1,0) and (2,0,1) after fitting a model with only ``basic''
        (single-nucleotide) mutations to sequence simulated under the model
        of Table~\ref{tab:cpg_results}, that also includes CpG hypermutability
        as well as $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$ mutations.
        Note that in the table, the underscore means ``any nucleotide''.
        \label{tab:resid1}
    }
\end{table}

%EM "These patterns": Is it exactly the patterns in the true model?
%EM Would we need to have prior information about CpGs to pick those additional terms?
%PR No - I've added explanation - see above.
After adding these patterns into the model and re-fitting, we searched for evidence of additional processes happening in the data.
To do so, we calculated length 2 residuals and (3,1,1) T-mer residuals.
Although some of the length 2 residuals had large $z$-scores,
none were as large as those seen in (3,1,1) T-mer residuals,
which are shown in Table~\ref{tab:resid2}.
These show an excess of $\nT\nC\nC \to \nT\nT\nC$ and $\nA\nG\nG \to \nA\nA\nG$ changes,
which were indeed the motifs used when simulating the data.
After adding these in,
% and initializing all parameters to 0.1
no further (3,1,1) T-mer residuals achieved
statistical significance after computing $p$-values with the Gaussian CDF
and applying a Bonferroni correction.
The final parameter values are shown in Table \ref{tab:cpg_results};
parameter estimates differ from the truth by less than 3\% (mostly within 1\%).

@@

\begin{figure}
    \begin{center}
        \includegraphics{writeup-plots/coverage_results}
    \end{center}
    \caption{
        Credible intervals for the four parameters of the dynamic Ising model,
        calculated using $(6,2,2)$ T-mers,
        from 61 MCMC runs on independently simulated sequences of $10^6$ bases
        differing at around 18.6\% of the sites.
        Dark bars show the 50\% credible intervals, and lighter grey bars the 95\% credible intervals.
        Values shown are relative parameters, obtained by dividing by the true value.
        \label{fig:ising_coverage}}
\end{figure}

@@

Nevertheless, we find significant effects for a number of biochemically-informed mutation features (Figure~\ref{fig:biochem_results}).
Specifically, we find strong CpG and UV damage effects \citep[][also seen by]{harris2015evidence}.
It may seem tenuous to look for mutational signatures associated with cancer in the germline,
or surprising that UV damage could play a significant role,
but it is reasonable that shared repair machinery might induce similar context-dependent errors.
Other likely sources of error in the germline include the effects of recombination
\citep{arbeithuber2015crossovers,myers2010drive}.

@@

This paper presents a tractable method
for doing statistical inference using observations of a particle model
in which probabilities of change are affected by nearby sites.
Such models appear in many fields, for instance,
as the dynamic counterparts to Markov random fields or spin glass models,
and models of DNA sequence evolution.

The methods described here provide a way to efficiently approximate full likelihoods,
which in many circumstances provide the optimal basis for inference \citep{neyman1933problem},
fast enough to allow Bayesian inference from Monte Carlo sampling.
The method is fast thanks to its dependence on only local neighborhoods
(so the time does not scale with sequence length once T-mer abundances are counted)
and its reliance on sparse matrices whose structure can be precomputed and cached.
It avoids approximations made by previous methods thanks to the use of T-mers
and explicit quantification of the necessary length of the ``arms'' of the T-mers.
We first discuss possible uses of the method,
followed by challenges (and possible solutions).

@@

\paragraph{Limitations}
The model presented assumes that the same process occurs at every location along the sequence,
and that the length of the sequence does not change (no insertions or deletions).
Neither of these assumptions are likely to be true for any data set of DNA sequence;
however, both assumptions are also commonly made in phylogenetics.
Some relaxation of the assumptions is possible -- for instance,
we can easily allow heterogeneity between sites by scaling rates by an overall gamma-distributed random factor \citep{yang1994maximum}.
Insertions and deletions are problematic, however.

Furthermore, we have not been able to put
quantitative bounds on how well we can approximate the full likelihood
with a given T-mer shape.
One reason this is difficult is that there are special cases of initial and final sequence
for which the likelihood is \emph{not} approximable by looking only at T-mer counts;
however, we believe such situations to be rare,
so that the full likelihood can be well approximated on a set of outcomes of high probability.
The TASEP model represents in some ways the worst-case of dependency propagation,
and has an analytic solution \citep{schutz1997exact}.
Futher theoretical work in this direction would be welcome.
In practice,
the approximation may be checked by computing the likelihood
using successively longer T-mers,
and checking for convergence as the length (of both base and overhang) increase.
